\input{chapter-header.tex}
% =============================================================================
\chapter{Conclusion}
\chaplabel{conclusion}
\minitoc
% =============================================================================
\introduction

In this chapter we summarize this dissertation.
We list the contributions, the published papers and the created software artifacts and their impact.

\begin{description}
\item[\chapref{background}] listed related work for this dissertation.
	We presented different types of metacircular high-level language \VMs.
	As a result we found only published results of two research \VM that have a unified model.
	Most other \VMs follow a clear separation between \VM-side and and isolated language-side.

\item[\chapref{reification}] presented basic tools required to facilitate the interaction with the low-level world.
	We presented a custom inspector framework which is now used by default in \PH.
	The second part explained how we use first-class object layouts to bridge the gap between high-level language-side and low-level memory.
	Both projects are crucial for metacircular \VM development and are direct results from the research conducted on the \P \VM.

\item[\chapref{benzo}] described a high-level low-level programming framework named \B.
	The core functionality of \B is to dynamically execute native-code generated at language-side.
	\B allows us to hoist typical \VM plugins to the language-side.
	Furthermore we show how code caching makes \B efficient and users essentially only pay a one-time overhead for generating the native code.
	
\item[\chapref{ffi}] presented a \NB, a stable foreign function interface (\FFI) implementation that is entirely written at language-side using \B.
	\NB is a real-world validation of \B as it combines both language-side flexibility with \VM-level performance.
	We show in detail how \NB outperforms other existing \FFI solutions on \PH.
	However, we observe that the underlying \B requires more effort to simplify low-level debugging and improve platform independence.

\item[\chapref{validation}] focused on two further \B applications: dynamic primitives and a language-side \JIT compiler.
	The dynamic primitives show how a metacircular infrastructure can be used dynamically and at runtime to reify part of the \VM necessary for compiling primitives.
	With the dynamic primitives we are able to alter a part of the execution semantics embedded into the language.
	This concept is taken further with the language-side \JIT presented in the second part of this chapter.
	With the \JIT we are able to control further aspects of the execution.
	We fully rely on language-side generated code instead of bytecodes which have tied interactions with the \VM.
	\VM-level \JIT, however, it is currently limited to simple expressions.
	Our \JIT shows that for certain applications a well-define interface with the low-level components of the \VM is required.

\item[\chapref{future}] summarized the limitations of \B and its application, furthermore we list undergoing efforts on the \B infrastructure and future work.
	We conclude that the \B approach for \VM interaction requires more support at \VM-level.
	Namely, missing dynamically accessible \VM-level reification make it hard to communicate with \VM internal components.
	Based on this observation we present a summary of another research \VM which also tries to overcome the limitations identified with the \B framework itself.
	Namely, the project is built around inspection and outside interaction during the complete development process.	
\end{description}

% =============================================================================
\section{Contributions}
% =============================================================================
The main contributions of this thesis are:
%
\begin{itemize}
	\item Description of the properties of an open and reflective language runtime.
	
	\item Implementation of \B, a dynamic high-level low-level programming framework for \PH.
	
	\item An in depth validation of \B with \NB, foreign function interface  implemented on top of \B.
	\NB proves the feasibility and efficiency of a dynamic high-level low-level programming framework.
	
	\item A \B-based language-side \JIT compiler showing the boundaries of \B.
	
	\item A road map for the future, bottom-up implementation of an open language runtime with full \VM-level reification using a platform independent \IR for low-level programming.
\end{itemize}


% =============================================================================
\section{Published Papers}
% =============================================================================

% -----------------------------------------------------------------------------
\subsection*{\href{http://dx.doi.org/10.1145/2048066.2048138}{Flexible object layouts: enabling lightweight language extensions by intercepting slot access}}
% -----------------------------------------------------------------------------

This paper contributed to \chapref{reification}.
Slots and Layouts as described in this paper are used in \PH 3.0 and newer.
The original implementation presented in this paper was implemented in an older \PH 1.0 image and was ported to \PH 3.0 in 2013.

\begin{description}
	\item[Abstract:] \emph{
Programming idioms, design patterns and application libraries often introduce cumbersome and repetitive boilerplate code to a software system.
Language extensions and external \textsc{dsl}s (domain specific languages) are sometimes introduced to reduce the need for boilerplate code, but they also complicate the system by introducing the need for language dialects and inter-language mediation.
To address this, we propose to extend the structural reflective model of the language with object layouts, layout scopes and slots.
Based on the new reflective language model we can 1) provide behavioral hooks to object layouts that are triggered when the fields of an object are accessed and 2) simplify the implementation of state-related language extensions such as stateful traits.
By doing this we show how many idiomatic use cases that normally require boilerplate code can be more effectively supported.
We present an implementation in \ST, and illustrate its usage through a series of extended examples.}

	\item[Authors:] Toon Verwaest, Camillo Bruni, David Gurtner, Adrian Lienhard and Oscar Nierstrasz. 
	\item[Revenue:] In Onward! 2011, Reno/Tahoe, Nevada, USA, 2011.
	\item[URL:] \url{http://dx.doi.org/10.1145/2048066.2048138}
\end{description}

% -----------------------------------------------------------------------------
\subsection*{\href{http://hal.inria.fr/hal-00840781}{Language-side Foreign Function Interfaces with \NB}}
% -----------------------------------------------------------------------------

This paper served as the basis for \chapref{ffi} and discusses the \B-based \FFI implementation \NB in detail.
\NB is used in production in the \PH 2.0 and newer.

\begin{description}
	\item[Abstract:] \emph{
		Foreign-Function-Interfaces (\FFIs) are a prerequisite for close system integration of a high-level language.
		With \FFIs the high-level environment interacts with low-level functions allowing for a unique combination of features.
		This duality has a strong impact on the implementation of the \FFI: it has to be flexible and fast at the same time.
		We propose \NB a language-side approach to \FFIs that only requires minimal changes to the \VM.
		\NB directly creates specific native code at language-side and thus combines the flexibility of a language-side library with the performance of a native plugin.}

	\item[Authors:] Camillo Bruni, Luc Fabresse, Stéphane Ducasse and Igor Stasenko. 
	\item[Revenue:] In International Workshop on \ST Technologies, Annecy, France, 2013.
	\item[URL:] \url{http://hal.inria.fr/hal-00840781}
\end{description}



% =============================================================================
\section{Software Artifacts}
% =============================================================================
During the work on this dissertation we produced several software artifacts.
Many projects emerged out of improving the infrastructure around the \PH development required to implement the core artifacts presented in this thesis.

\begin{description}
	\item[Collaboration on First-class Layouts and Slots:]
		In a collaboration with To\-on Verwaest (\SCG, Switzerland) we built a first implementation of first-class layouts and slots in a \ST system \cite{Verw11a}.
		In Collaboration with Martin Días (\RMoD, \INRIA) this initial version was ported to \PH and is now used in the current release candidate \urlfootnote{\PH 3.0}{http://files.pharo.org/image/30}.

	\item[\AsmJIT 64-bit Assembler:]
		To reuse the original research compilation pipeline built with \P \cite{Verw10a, Brun11a} a 64-bit extension was necessary to the initial \urlfootnote{\AsmJIT implementation for \PH}{http://smalltalkhub.com/\#!/~Pharo/AsmJit}.
		The extension is included in the current stable \urlfootnote{\PH release 2.0}{http://files.pharo.org/image/20}.

	\item[Collaboration on the \WF Dynamic Primitives:]
		We collaborated on Guido Chari's (\UBA, Argentina) \WF Dynamic Primitive compiler, which resulted in paper currently under submission\cite{Char13a}.
		The implementation is a prototype and is not used in production.

	\item[Collaboration on the Mate \VM Prototype:]
		In collaboration with Guido Chari (\UBA, Argentina), Javier Pímas (\UBA, Argentina) and Clement Bera (\RMoD, \INRIA) several stages of a prototype \VM were built.
		The implementation mainly follows the concept of an dynamic language runtime which controls every aspect at language-side.
		The current language runtime is in a early prototype phase that allows us to explore new \VM and language concepts, however it is not production ready.
		Guido Chari will further explore new concepts of Mate in his Ph.D.
		Clement Bera, after finishing his engineering contract at \RMoD, will continue to work as a Ph.D. on the same system.

	\item[\VirtualCPU Compilation Toolchain:]
		In collaboration with Clement Bera (\RMoD, \INRIA) and Igor Stasenko (\RMoD, \INRIA) we built a prototype compilation toolchain based on the original work of Pinocchio.
		The current implementation is a working prototype.
		Plans exist to integrate a streamlined version into \PH to server as a platform independent backend to our \B-based \FFI implementation used in \PH.

	\item[\NBJ Language-side \JIT Compiler:]
		As a third case study for the \B framework we implemented a language-side \JIT compiler. 
		The current implementation is a prototype that is capable of directly transforming simple methods to executable code.
		Unlike its \VM-level counterpart it is based as a simple visitor over the intermediate bytecode format already present at language-side.

	\item[Inspector Framework for \PH:] 
		An important part of reifying concepts is the possibility to inspect and manipulate these objects.
		We wrote a new inspector framework which is used in the latest \PH release.
		It allows to quickly define new views on domain objects, an indispensable requirement for interacting with complex data objects.
		Next to the everyday usage in \PH it is actively used for the Mate \VM prototype where we need transparent access to internal structures of the \VM.

	\item[Command Line Test Interface for \PH:]
		In order to perform continuous integration in a maintainable fashion we developed a new modular command line interface for \PH. 
		It is used in production on the \urlfootnote{\PH build server}{http://ci.inria.fr/pharo/} alongside with simple \urlfootnote{installer scripts}{http://get.pharo.org/}.
		
	\item[Validation Framework:]
		In order to improve the integration life cycle of \PH we developed a validation framework.
		Many changes in the core \PH were required to support the previously presented tools.
		To encourage faster integration we validate each proposed change by running lint rules and all unit tests.
		The tool generates a validation report as separate website and interacts with the issue tracker.
\end{description}


% =============================================================================
\section{Impact of the Thesis}
% =============================================================================
Many engineering artifacts built during this dissertation are used in production for \PH.
The inspector framework allowed us to create many specific views on common objects in \PH, improving the development in \PH.
For instance the inspectors are tightly integrated into the debugger.
Outside the image, we have contributed to improve the \PH continuous integration work-flow.
Identifying the limitations of the \B framework will trigger more specific development efforts to improve the debugging capabilities.

Our dissertation helped to bootstrap the \MATE research \VM which eventually will cope with the limitations of the \B approach.


% =============================================================================
\input{chapter-footer.tex}