\input{chapter-header.tex}
% =============================================================================

\chapter{Conclusion}
\chaplabel{conclusion}
\minitoc

% =============================================================================

% =============================================================================
\section{Contributions}
% =============================================================================
% detection of VM level reflectivity => reification?
% identification of tools that cross vm boundaries
% identification of partial solution of these tools using benzo
% identification of the restrictions of a benzo-based tool
The main contributions of this thesis are:
\begin{enumerate}
	\item Description of the properties of an open and reflective language runtime.
	\item Implementation of 3 main software artifacts to show the feasibility of an open VM.
	\item A roadmap for the future, bottom-up implementation of an open language runtime.
\end{enumerate}

% =============================================================================
\section{Published Papers}
% =============================================================================
The following papers have been published and accepted:
\begin{itemize}
	\item Toon Verwaest, Camillo Bruni, David Gurtner, Adrian Lienhard and Oscar Nierstrasz. 
\emph{Flexible object layouts: enabling lightweight language extensions by intercepting slot access}. 
In Onward! 2011, Reno/Tahoe, Nevada, USA, 2011.
	\item Camillo Bruni, Luc Fabresse, Stéphane Ducasse and Igor Stasenko. 
\emph{Lan\-guage-side Foreign Function Interfaces with NativeBoost}. 
In International Workshop on Smalltalk Technologies, Annecy, France, 2013.
\end{itemize}


% =============================================================================
\section{Software Artifacts}
% =============================================================================
% 64bit ASM implementation
% Collaborated work on basic MATE prototype
% Collaborated work on Waterfall
% Collaborated work on Nabujito
% Cleaned up VM builds for easier contributions and faster prototyping

\begin{description}
	\item[Collaboration on First-class Layouts and Slots:]
In a collaboration with Toon Verwaest (SCG, Switzerland) we built a first implementation of first-class layouts and slots in a Smalltalk system \cite{Verw11a}.
In Collaboration with Martin Días (RMoD, INRIA) this initial version was ported to Pharo\footnote{\url{}}and is now used in the current release candidate Pharo 3.0\footnote{\url{http://files.pharo.org/image/30}}.

	\item[AsmJIT 64-bit Assembler:]
To reuse the original research compilation pipeline built with Pinocchio \cite{Verw10a, Brun11a} a 64-bit extension was necessary to the initial AsmJit implementation for Pharo\footnote{\url{http://smalltalkhub.com/\#!/~Pharo/AsmJit}}.
The extension is included in the current stable Pharo release 2.0\footnote{\url{http://files.pharo.org/image/20}}.

	\item[Collaboration on the Waterfall Dynamic Primitives:]
We collaborated on the Waterfall Dynamic Primitive compiler together with Guido Chari (UBA, Argentina), which resulted in paper currently under submission\cite{Char13a}.
The implementation is a prototype and is not used in production.

	\item[Collaboration on the Mate VM Prototype:]
In collaboration with Guido Chari (UBA, Argentina), Javier Pímas (UBA, Argentina) and Clement Bera (RMoD, INRIA) several stages of a prototype VM were built.
The implementation mainly follows the concept of an dynamic language runtime which controls every aspect at language-side.
The current language runtime is in a early prototype phase that allows us to explore new VM and language concepts, however it is not production ready.
Guido Chari will further explore new concepts of Mate in his PhD.
Clement Bera, after finishing his engineering contract at RMoD, will continue to work as a PhD on the same system.

	\item[VirtualCPU Compilation Toolchain:]
In collaboration with Clement Bera (RMoD, INRIA) and Igor Stasenko (RMoD, INRIA) we built a prototype compilation toolchain based on the original work of Pinocchio.
The current implementation is a working prototype.
Plans exist to integrate a streamlined version into Pharo to server as a platform independent backend to our Benzo-based FFI implementation used in Pharo.

	\item[Nabujito Language-side JIT Compiler:]
As a third case study for the Benzo framework we implemented a language-side JIT compiler. 
The current implementation is a prototype that is capable of directly transforming simple methods to executable code.
Unlike its VM-level counterpart it is based as a simple visitor over the intermediate bytecode format already present at language-side.

	\item[Inspector Framework for Pharo:] 
An important part of reifying concepts is the possibility to inspect and manipulate these objects.
We wrote a new inspector framework which is used in the latest Pharo release.
It allows to quickly define new views on domain objects, an indispensable requirement for interacting with complex data objects.
Next to the everyday usage in Pharo it is actively used for the Mate VM prototype where we need transparent access to internal structures of the VM.

	\item[Commandline Test Interface for Pharo:]
In order to perform continuous integration in a maintainable fashion we developed a new modular command line interface for Pharo. 
It is used in production on the Pharo build server\footnote{\url{http://ci.inria.fr/pharo}} alongside with simple installer scripts\footnote{\url{http://get.pharo.org}}.
\end{description}


% =============================================================================
\section{Future Work}
% =============================================================================

% =============================================================================
\input{chapter-footer.tex}